---
title: "Multivariate Data Analysis: Workshop1"
output: pdf_document
date: "2024-01-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading a dataset

We will work on the (famous) Titanic dataset. Load the Titanic data set from a csv file (note that dots were used to represent missing data)

```{r}
Titanic <- read.csv("Titanic.csv",na.strings=".")
#
# Define Sex as factor (0 = Female, 1 = Male)
#
Titanic$Sex <- factor(Titanic$Sex,levels=c(0,1),labels=c("Female","Male"))
```

## Simple computations

### Table

We can represent the empirical joint distribution table (contingency table, cross tab, 2-way table) containing the counts of each categories.

```{r}
table(Titanic$Sex,Titanic$Survived)
```

We can transform this into proportion table:

```{r}
table(Titanic$Sex,Titanic$Survived)/nrow(Titanic)
```

The function \texttt{ftable} does the same thing but with a "flat" matrix.

### Marginal Distributions

The empirical marginal distributions are obtained by considering only the column of interest (i.e. the table contain a sample from (X,Y), so the first column is a sample from X). We can compute, run tests and infer on these samples as usually in the univariate case.

```{r}
table(Titanic$Survived)/nrow(Titanic) #proportion of survivants
```

It is also possible to add margins to a contingency table:


```{r}
tableTitanic <- table(Titanic$Sex,Titanic$Survived)/nrow(Titanic)
addmargins(tableTitanic)
```

## Conditional distributions

Now we want to understand the dependencies between the observations. The conditional density between $X$ and $Y$ writes:

\[ f_{X\mid Y}(x \mid y) = \frac{f_{X,Y}(x,y)}{f_Y(y)} \]

Ex: How do you estimate the probability of a woman to survive?

By using the help, find how to use \texttt{prop.table} to do these computations.

## Independence tests

By using \texttt{summary} on a table, you can perform independence $\chi^2$ tests:

```{r}
summary(table(Titanic$Sex,Titanic$Survived))
```

Remind yourself how to read this result.

## Some graphical representations

A simple way to graphically represent a table is to produce a \emph{mosaic}:

```{r}

plot(tableTitanic,main="Titanic",ylab="Survived",xlab="Sex",col=rainbow(2))

```

With a larger dataset, you can envision a larger spectrum of colors, for example using the \texttt{HairEyeColor} dataset.

```{r}

a <- as.table(HairEyeColor[,,"Male"]) # Extract table for Male category
a

```


```{r}

plot(a,main="Hair and eye colour (Male)",ylab="Eye",xlab="Hair",col=rainbow(4))

```

Ex. Do the same with \texttt{b} which will contain the same results for the Female category.

You can represent all the results of this bivariate categorical distribution on a grouped barplot:

```{r}

barplot(t(a),main="Hair and eye colour (Male)",col=rainbow(4),beside=T,xlab="Hair color",
legend=T,args.legend=list(x="topleft",bty="n"))

```

### More than two variates

Use the \texttt{ftable} function on the \texttt{HairEyeColor} dataset, and understand it.

Load the \texttt{Personal.csv} dataset it contains 4 variables, check how to use the \texttt{row.vars} and \texttt{col.vars} arguments to change the representation of the table.

## Numerical variables

In the case of continuous variables, covariance and correlation are better way to understand relationships between variables:

\[ S_{x,y} = \frac{\sum_i (x_i - \bar{x} )(y_i - \bar{y} )}{n}, \quad r_{x,y} = \frac{S_{x,y}}{S_x S_y}. \]

Use the \texttt{airquality} dataset, considering the first 4 variates to be continuous. Check the help on this dataset.

```{r}
air=airquality[,1:4]
```

There is missing data on this dataset. Check how the different functions react to the way to treat the missing data. For example:

```{r}

colMeans(air)

colMeans(air,na.rm=T)

```

Use the \texttt{apply} function to compute empirical variances of the variables.

You can use the \texttt{cor} function to plot correlation matrix, the \texttt{cov} function to plot the covariance matrix and the \texttt{cov2cor} function to convert a covariance matrix into a correlation matrix.


### Visualisation of the relationships

First, you can produce simple scatterplot of two variables.

```{r}
plot(air$Temp,air$Wind,ylab="Wind",xlab="Temperature (°F)")
```

You can then fitt a Loess curve to this plot:

```{r}

scatter.smooth(air$Temp,air$Wind,ylab="Wind",xlab="Temperature (°F)",lpars=list(col="red"))

```

You can visualise all the pairs of variables at once using:

```{r}

plot(air) #same as pairs()

```

For three variables, you can use the less readable 3D scatterplots:

```{r}

#install.packages("scatterplot3d")
library(scatterplot3d)
scatterplot3d(air$Temp,air$Wind,air$Solar.R,color="blue")

```

Ex. Do the same with the \texttt{iris} dataset.